Notes

### 2020-01-22 Notes/Notes.tex
Updated notes

\author{}

@@

>>>>

\author{C. C. Chang, C. C. Chen, C. K\"orber, T. Humble, J. Ostrowski}

@@

\section{Minimum Dominating Set Problem}
We want to find a Minimum~Dominating~Set (MDS) for a given graph $G(V,E)$ (defined by it's vertices $V$ and edges $E$).

\begin{itemize}
\item A \textit{dominating~set} $D$ is a set of vertices $V$ such that all vertices in the graph which are not in $D$ are adjacent to at least one point in this set.
In other words: each vertex $i \in V$ is either in the dominating set $i \in D$ or in the neighborhood of a vertex in the dominating set $i \in \mathcal N_j$ for $j \in D$ (see also eq.~\eqref{eq:ms-constraint}).

\item A \textit{minimal~dominating~set} is a dominating set which cannot be reduced by taking out vertices---if you are able to remove a vertex from your dominating set and still have a dominating set, it was not minimal.

\item Finally, a \textit{minimum~dominating~set} $D_\text{min}$ (note that minimum $\neq$  minimal) is the smallest minimal dominating set you can find.
Thus a Minimum~Dominating~Set is a minimal~dominating~set but not necessarily the other way around.
Note that there can be more then just one minimum~dominating~set for a given graph.

\item The \textit{Dominating~Set~Problem} (DSP) asks whether the number of vertices in the the smallest dominating set $\gamma(G) \equiv |D_\text{min}|$ is smaller or equal to a given integer number $\gamma(G) \leq K$.
If one was able to find a global MDS, one would solve this problem for a given graph.
\end{itemize}

The integer linear programming representation of the MDS problem is given by the minimization condition
\begin{equation}\label{mds:min-con-classical}
	\min_{i \in V}\left( \sum_i x_i \right)\, ,
\end{equation}
and the $|V|$ (number of all vertices) constraint conditions
\begin{equation}\label{eq:ms-constraint}
	x_i + \sum_{j \in \mathcal N_i} x_j \geq 1 \, , \qquad x_i \in \{ 0, 1\} \, , \qquad \forall_{i \in V} \, ,
\end{equation}
where $x_i$ represents if vertex $i$ is in the dominating set ($i\in D_\text{min} \Rightarrow x_i = 1$) and $\mathcal N_i$ is the neighborhood of vertex $i$ (not including $i$ itself).

To allow the implementation of inequalities on D-Wave, we first map eq.~\eqref{eq:ms-constraint} to it's slack variable representation
\begin{equation}\label{mds:con-slack-classical}
	x_i  - s_{i} - 1 + \sum_{j \in \mathcal{N}_i} x_j = 0 \, , \quad 0 \leq s_{i} \leq |\mathcal{N}_i| \, .
\end{equation}
The solution to the DSP is found by finding the optimal vector $\vec \psi = (\vec x, \vec s)^T$ such that the objective function $f$ is minimal
\begin{equation}\label{mds:objective-function}
	f(\vec x, \vec s)
	=
	\sum_{i\in V} x_i
	+ p \sum_{i \in V} \left(x_i  - s_{i} - 1  + \sum_{j \in\mathcal{N}_i} x_j\right)^2
	\, ,
	\qquad
	p \geq |V| + 1 \, .
\end{equation}
To guarantee that the global minimum solution fulfills the constraints, $p$ needs to be larger than the minimization condition eq.~\eqref{mds:min-con-classical}.
In principal, for any graph $p = |V| + 1$ is sufficient---nevertheless it might be that one has to optimize this parameter in order to simplify convergence on the D-Wave.

The values of the vertex variables $x_i$ are already in a bit representation and thus can be directly mapped onto QUBO components.
The slack variables can be integers and thus have to be mapped onto bits.
To ensure that eq.~\eqref{mds:con-slack-classical} is fulfilled, we introduce the following mapping procedure:
For a given slack variable $s_i$, we find the smallest integer $b_i$ such that $2^{b_i}$ is larger than the number of neighbors of vertex $i$
\begin{equation}
	|\mathcal{N}_i| \leq 2^{b_i} - 1 = \max(s_i)
	\quad \Rightarrow \quad 
	s_i = \sum_{k=0}^{b_i-1} 2^k q_{ik}
	\, , \qquad 
	q_{ik} = 0, 1
	\, \qquad
	\forall_{i \in V}
	\, .
\end{equation}
For this choice, the objective function can be written in terms of a vector-matrix-vector product of QUBO $Q$ 
\begin{align}\label{mds:qubo}
	Q &= 
	\begin{pmatrix}\Omega & 0 \\ 0 & 0 \end{pmatrix}
	+ p
	\begin{pmatrix}\alpha & \beta \\ 0 & \gamma \end{pmatrix}
	\, , & 
	f(\vec x, \vec s)
	&=
	\vec \psi \cdot Q \cdot \vec \psi + p |V|
	\, . 
\end{align}
The bit vector $\vec q$ is related to the slack variable $\vec s$ by the bitmap $\vec B$
\begin{equation}
	\vec \psi \to \begin{pmatrix}
		\vec x \\ \vec q
	\end{pmatrix}
	\, , \qquad
	\vec q = (q_{00}, q_{01}, \cdots q_{0 b_0}, q_{10}, q_{11}, \cdots )^T
	\, , \qquad
	\vec s = B \vec q
	\, .
\end{equation}
The first matrix in eq.~\eqref{mds:qubo} represents the minimization condition eq.~\eqref{mds:min-con-classical} while the second matrix describes the constraint eq.~\eqref{eq:ms-constraint}.
The derivation of the QUBO utilizes that components of the bit vector squared are equal to the component, $\psi_\nu^2 = \psi_\nu$, and thus the quadratic QUBO element is able to represent the original objective function $f$ modulo a known constant.

The block matrices are determined by the linear and quadratic components of eq.~\eqref{mds:objective-function}
\begin{align}
	\vec x \cdot \Omega \cdot \vec x 
	&=
	\sum_{i \in V} x_i
	\, , \\
    \vec x \cdot \alpha \cdot \vec x
    &= 
    \sum_{i \in V} \left[
    	x_i^2 + 2 x_i \sum_j x_j  - 2 x_i - 2 \sum_j x_j + \left(\sum_j x_j\right)^2
	\right]
    \, , \\
    \vec s \cdot \beta \cdot \vec x
    &=
    \sum_{i \in V} \left[ - 2 x_i s_i - 2 s_i \sum_j x_j\right]
    \, , \\
    \vec s \cdot \gamma \cdot \vec s 
    &= 
    \sum_{i \in V} \left[s_i^2 + 2 s_i\right] \, .
\end{align}
Given the adjacency matrix $A$\footnote{The adjacency matrix used in this context is defined as the symmetric matrix representing the graph with and zero diagonal entries. Other entries are either equal to one or zero.} and the slack-bitmap $B$, one finds
\begin{align}
	\Omega &= \mathbbm{1} \, ,
    &
    \alpha &= - \mathbbm{1} + 2 A - 2 \text{diag}(|A|) + A^2
    \, , \\
    \beta &= -2 (\mathbbm{1} + A) B
    \, , &
     \gamma &= B^T B + 2 \text{diag}(|B|)
    \, .
\end{align}
The result of the matrix operation $|M|$ defines a vector given by the sum over the matrix columns $|M| = \sum_{ij} M_{ij} \vec e_j$.

\subsection{Remarks on Scaling}
The QUBO $Q$ is upper block triangular (does this word exist?), $\gamma$ is block diagonal and $\alpha, \beta$ are relatively dense because of the neighbors-to-neighbors entries ($A^2$) and neighbors-to-slack-bits entries ($A B$).
Assuming there is maximal number of edges per vertex which does not scale with the number of vertices $|V|$, the total scaling is
\begin{equation}
	\text{rank}(Q) \sim |V| \, ,
\end{equation}
if the maximal number of edges is in some way proportional to $|V|$ one finds
\begin{equation}
	\text{rank}(Q) \lesssim |V| \log(|V|) \, .
\end{equation}

\subsection{Examples}
I have implemented a Python module which takes a graph (defined by a set of edges) as input, computes the matrices $A$ and $B$, and eventually computes the QUBO $Q$.
Next, the module classically searches the full bit-space for small example graphs to verify that the solutions returned by the algorithm make sense.
You can find a few example plots in fig.~\ref{fig:mds-sol-1}.

If you run `\texttt{python qlp/mds/qubo.py}'\footnote{You have to install the scripts by running `\texttt{python~-m~pip~install~.}' before you are able to run the code.} from the repo root, you will generate a random graph, compute the minimum QUBO, find the global minimum solutions and visualize them.

\begin{figure}[htb]
	\centering
	\includegraphics[width=0.64\textwidth]{mds-sol-n5-s.png}
	\includegraphics[width=0.32\textwidth]{mds-sol-n5.png}
	\includegraphics[width=0.49\textwidth]{mds-sol-n6.png}
	\includegraphics[width=0.49\textwidth]{mds-sol-n7.png}
	\caption{
		\label{fig:mds-sol-1} Vertex representation for global minimum solution of $\vec \psi \cdot Q \cdot \vec \psi$ for random 5-, 6- and 7-vertex graphs.
		Green vertices are included in the MDS.
	}
\end{figure}
